Relational Parametricity for Algebraically Indexed Types

### 2012-06-12 paper/section-general-framework.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

\fixme{Update this introduction}

We now develop a general framework for discussing
algebraically-indexed types and their relational interpretations. We
have chosen a particular order for our presentation: presenting the
semantics of types before we present the syntax of programs and their
typing relation. We have deliberately selected this order to emphasise
the importance of establishing what types \emph{mean} over the
language used to write programs.

We first present, in \autoref{sec:algebraically-indexed-types}, the
syntax of the language of algebraically-indexed types. Our general
framework is parameterised by a choice of multi-sorted algebraic
theory to be used for indexing types, and the collections of primitive
types and
operations. %FIXME: refer forwards to the names of the parameters
We define substitution of index terms, and what it means for two types
to be equal. Equality on types is generated from the chosen
multi-sorted algebraic theory.

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: the
\emph{index-erasure} interpretation. The index-erasure interpretation
provides an ``underlying'' semantics to each type (and term), which we
think of as an abstraction of the actual operational behaviour of a
term. By erasing all the index terms, we can immediately see that the
run-time behaviour of programs cannot depend on index terms. We then
define the \emph{relational interpretation}, which interprets every
type as a binary relation on its index-erasure interpretation. The
relational interpretation acts as ``commentary'' on the underlying
interpretation, constraining the index-erasure interpretations that
are allowed. In particular, we use the relational interpretation to
enforce the invariance properties that we identified in
\autoref{sec:motivating-examples}.

Throughout this section, we use Kennedy's original units of measure
system as a running example in order to provide a concrete example of
our general framework, and also to demonstrate that we have in fact
generalised Kennedy's original development.

@@

The indexing terms and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We will use the
  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
  arbitrary sorts taken from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$ (we use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$).
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
  of the arguments of each primitive type.
\end{enumerate}

\begin{example}[Origin Invariance]
  \fixme{Do this, then update the units of measure example}
\end{example}

\begin{example}[Units of Measure]
  To instantiate our general framework as Kennedy's original units
  of measure system, we would take the following settings. For the
  collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
  the collection of index operations, we take $\mathit{IndexOp} = \{1,
  -\cdot-, -^{-1}\}$, with:
  \begin{eqnarray*}
    \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
    \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
    \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
  \end{eqnarray*}
  The intended interpretation of these operations is group unit, group
  multiplication and group inverse, respectively. When we discuss
  equational theories for index expressions in
  \autoref{sec:type-equality} below, we will impose the abelian group
  laws on these operations.

  The primitive types for the units of measure example are taken to be
  $\PrimTypeSet = \{ \mathsf{num} \}$, with
  $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
\end{example}

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index terms $\Delta \vdash e : s$
and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index terms are particularly simple: either an index term is a
variable that appears in the context (rule \TirName{IVar}), or it is
an application of an index operation taken from $\IndexOpSet$ to other
index terms (rule \TirName{IOp}). The rules for well-indexed types
include the usual rules for constructing types of the simply-typed
$\lambda$-calculus with unit, sum and tuple types (rules
\TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index terms
$e_1,...,e_n$, the well-indexed type $\tyPrim{X}{e_1,...,e_n}$. The
rule \TirName{TyForall} permits the formation of universally
quantified types, where the universal quantification ranges over all
index terms of some sort.

\subsubsection{Simultaneous Substitutions of Index Terms}
\label{sec:simultaneous-substitution}

It will be technically convenient to express substitution of index
terms in our framework in terms of simultaneous substitutions.  Given
a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1,
..., i_n \mathord: s_n$, a \emph{simultaneous substitution} $\Delta
\vdash \sigma \Rightarrow \Delta'$ is a sequence of terms $\sigma =
(e_1,...,e_n)$ such that for each $j$ such that $1 \leq j \leq n$,
$\Delta \vdash e_j : s_j$. Given a simultaneous substitution $\Delta
\vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a variable $i_j
\mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for the index term
$e_j$.

We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\Delta'$. When we discuss the relational interpretation of types
below, it will be useful to think of any sequence of sorts as an index
context. \fixme{flesh this out}

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index terms is preserved by addition of extra items
to the context.

@@

\subsubsection{Index Term Equality and Type Equality}
\label{sec:type-equality}

\fixme{Consider updating this example} Much of the power of indexing
types by the terms of an algebraic theory comes from the equations of
the theory, and the fact that these equations are inherited by the
types. In the units of measure example, the well-indexed types
$\tyPrim{X}{u_1\cdot u_2}$ and $\tyPrim{X}{u_2 \cdot u_1}$ are
considered equal by the type system, due to the commutativity of the
$-\cdot-$ operation in the theory of abelian groups.

@@

Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
equality between index terms $\Delta \vdash e \equiv e' : s$ by the
following rules, which allow us to use substitution instances of
axioms, and ensure that judgmental equality is a congruence relation
with respect to the index operations in $\IndexOpSet$:
\begin{mathpar}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}

@@

\begin{example}[Origin Invariance]
  FIXME
\end{example}

\begin{example}[Units of Measure]
  For the units of measure system, seen as an instantiation of our
  general framework, the set $\IndexAxiomSet$ consists of the abelian
  group axioms:
  \begin{displaymath}
    \begin{array}{l}
      u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
      u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
      u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
      u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
    \end{array}
  \end{displaymath}
\end{example}

The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
terms generates the judgmental equality $\Delta \vdash A \equiv B
\isType$ on types. The basic rule generating judgmental equality on
types states that two applications of primitive types are equal if
their index term arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for judgmental equality on types ensure that it
is a congruence relation on types, and that it is an equivalence
relation (i.e.~judgmental equality on types is reflexive, symmetric
and transitive). For example, for universally quantified types we have
the following congruence rule:
\begin{displaymath}
  \inferrule*
  {\Delta, i : s \vdash A \equiv B \isType}
  {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
\end{displaymath}
The congruence rules for the other type formers are similar.

The judgmental equalities on index terms and on types are preserved by
substitution:
\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
    \sigma^*e \equiv \sigma^*e' : s$; and
  \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
    \sigma^*A \equiv \sigma^* B \isType$.
  \end{enumerate}
\end{lemma}

We define a pair of simultaneous substitutions $\sigma, \sigma' :
\Delta \Rightarrow \Delta'$ to be judgmentally equal if their
component terms are judgmentally equal in the context $\Delta$: i.e.~
if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
simultaneous substitutions are judgmentally equal.

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index terms $e_1,...,e_n$. Accordingly, we assume that each primitive
type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0.2em}}c@{\hspace{0em}}c}
    \begin{array}{l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretation of the
universal quantifier ignores the indexing: the interpretation of the
type $\forall i\mathord:s.A$ is exactly the interpretation of the type
$A$.

\begin{example}[Origin Invariance]
  \fixme{do this}
\end{example}

\begin{example}[Units of Measure] Kennedy's original units of measure
  system seen as an instantiation of our system uses the assignment
  $\tyPrimSem{\mathsf{unit}} \isDefinedAs \mathbb{Q}$, where
  $\mathbb{Q}$ is the set of rational numbers.

  FIXME: Also do the primitive operations
\end{example}

@@

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define. \fixme{Backwards ref}

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index term equality: for any
$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
$\Delta \vdash \sigma \Rightarrow \primTyArity(\tyPrimNm{X})$ and
$\Delta \vdash \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$ such
that $\Delta \vdash \sigma \equiv \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\ \sigma = \rho\
\tyPrimNm{X}\ \sigma'$.

@@

The following lemma states that the relation interpretation of types
that we have defined in this section behaves well: the first part of
the lemma states that two types that are judgmentally equal are given
equal relational interpretations, and the second part states that
substitution of index terms in types can be interpreted by the
composition of relational environments with simultaneous
substitutions. Note that both parts of the lemma depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and $\sigma : \Delta \Rightarrow
    \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}

@@

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework. In the \fixme{next section} we will
use this theorem to derive several interesting results about the
programs expressible in instances of our general framework.

@@

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

>>>>

We now present a general framework for discussing
algebraically-indexed types and representation independence. We define
the syntax of algebraically-indexed types, then their semantics and
finally we present a general programming language for
algebraically-indexed types, with an abstraction theorem.

@@

The index expressions and types of an instantiation of our general
framework are derived from the following data:
\begin{enumerate}
\item A collection $\SortSet$ of index sorts. We will use the
  meta-syntactic variables $s,s_1,s_2,...$ and so on to stand for
  arbitrary sorts taken from $\SortSet$.
\item A collection $\IndexOpSet$ of index operations, with a function
  $\indexOpArity : \IndexOpSet \to \SortSet^* \times \SortSet$ (we use
  the notation $A^*$ to denote the set of lists of elements of some
  set $A$).
\item A collection $\PrimTypeSet$ of primitive types, with a function
  $\primTyArity : \PrimTypeSet \to \SortSet^*$, describing the sorts
  of the arguments of each primitive type.
\end{enumerate}

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl_2,
  \SynGL_2, \SynOrth_2, \SynGL_1 \}$. For the index operations, we
  have additive group structure on $\SynTransl_2$, multiplicative
  group structure on $\SynGL_1$, $\SynGL_2$ and $\SynOrth_2$,
  injections from $\SynOrth_2$ and $\SynGL_1$ into $\SynGL_2$,
  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
  $G \in \{\SynGL_1, \SynGL_2, \SynOrth_1\}$, with the following
  assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl_2) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl_2, \SynTransl_2], \SynTransl_2) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl_2], \SynTransl_2) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth_2], \SynGL_2) &
      \indexOpArity(\iota_1) & ([\SynGL_1], \SynGL_2) \\
      \indexOpArity(\det) & ([\SynGL_2], \SynGL_1) &
      \indexOpArity(|\cdot|) & ([\SynGL_1], \SynGL_1)
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL_2, \SynTransl_2]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL_1]$.
\end{example}

% \begin{example}[Logical Information Flow]
%   \fixme{Do this}
% \end{example}

% \begin{example}[Units of Measure]
%   To instantiate our general framework as Kennedy's original units
%   of measure system, we would take the following settings. For the
%   collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
%   the collection of index operations, we take $\mathit{IndexOp} = \{1,
%   -\cdot-, -^{-1}\}$, with:
%   \begin{eqnarray*}
%     \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
%     \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
%     \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
%   \end{eqnarray*}
%   The intended interpretation of these operations is group unit, group
%   multiplication and group inverse, respectively. When we discuss
%   equational theories for index expressions in
%   \autoref{sec:type-equality} below, we will impose the abelian group
%   laws on these operations.

%   The primitive types for the units of measure example are taken to be
%   $\PrimTypeSet = \{ \mathsf{num} \}$, with
%   $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
% \end{example}

@@

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some sort.

\subsubsection{Simultaneous Substitution of Index Expressions}
\label{sec:simultaneous-substitution}

It will be technically convenient to express substitution of index
expressions in our framework in terms of simultaneous substitutions.
Given a pair of index contexts $\Delta$ and $\Delta' = i_1 \mathord:
s_1, ..., i_n \mathord: s_n$, a \emph{simultaneous substitution}
$\Delta \vdash \sigma \Rightarrow \Delta'$ is a sequence of
expressions $\sigma = (e_1,...,e_n)$ such that for each $1 \leq j \leq
n$, $\Delta \vdash e_j : s_j$. Given a simultaneous substitution
$\Delta \vdash \sigma = (e_1,...,e_n) \Rightarrow \Delta'$ and a
variable $i_j \mathord: s_j$ in $\Delta'$, we write $\sigma(i_j)$ for
the index expression $e_j$.

It will be useful to think of any sequence of sorts as an index
context. In particular, we will make use of simultaneous substitutions
of the form $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, since these are exactly sequences of
index arguments suitable for the primitive type $\tyPrimNm{X}$.

We write $\Delta \Rightarrow \Delta'$ for the set of all simultaneous
substitutions $\sigma$ such that $\Delta \vdash \sigma \Rightarrow
\Delta'$. The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is just the sequence of variables in
$\Delta$: $\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. We will also define composition
of simultaneous substitutions below, but first we must define the
application of a simultaneous substitutions to index expressions.

@@

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$ where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$ and a
variable/type pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, then we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. Note that we have implicity used the fact that
well-sortedness of index expressions is preserved by addition of extra
items to the context.

\subsubsection{Index Expression Equality and Type Equality}
\label{sec:type-equality}

Much of the power of indexing types by the expressions of an algebraic
theory comes from the equations of the theory. In the two-dimensional
geometry example, the types $\tyPrim{vec}{B,t_1 + t_2}$ and
$\tyPrim{vec}{B, t_2 + t_1}$ are considered equal by the type system,
due to the commutativity of the $+$ operation.

@@

Given a set $\IndexAxiomSet$ of axioms, we generate the judgmental
equality between index expressions $\Delta \vdash e \equiv e' : s$ by
the following rules, which allow us to use substitution instances of
axioms, and ensure that judgmental equality is a congruence relation
with respect to the index operations in $\IndexOpSet$:
\begin{mathpar}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}

@@

\begin{example}[Two-Dimensional Geometry]
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl_2 \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl_2 \\
      t_1, t_2, t_3 : \SynTransl_2 \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl_2 \\
      t : \SynTransl_2 \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl_2 \\
      t_1, t_2 : \SynTransl_2 \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl_2 \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL_1$ forms an abelian
  group under multiplication, and the sorts $\SynGL_2$ and
  $\SynOrth_2$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL_2$. The absolute value of the
  determinant of orthgonal transformations is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

% \begin{example}[Logical Information Flow]
%   \fixme{Do this}
% \end{example}

% \begin{example}[Units of Measure]
%   For the units of measure system, seen as an instantiation of our
%   general framework, the set $\IndexAxiomSet$ consists of the abelian
%   group axioms:
%   \begin{displaymath}
%     \begin{array}{l}
%       u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
%       u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
%       u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
%       u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
%     \end{array}
%   \end{displaymath}
% \end{example}

The judgmental equality $\Delta \vdash e \equiv e' : s$ on index
expressions generates the judgmental equality $\Delta \vdash A \equiv
B \isType$ on types. The basic rule generating judgmental equality on
types states that two applications of primitive types are equal if
their index expression arguments are equal:
\begin{displaymath}
  \inferrule*
  {\{ \Delta \vdash e_j \equiv e'_j : s_j\}_{1\leq j \leq n}}
  {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,...,e'_n} \isType}
\end{displaymath}
The rest of the rules for judgmental equality on types ensure that it
is a congruence relation on types, and that it is an equivalence
relation (i.e.~judgmental equality on types is reflexive, symmetric
and transitive).
% For example, for universally quantified types we have
% the following congruence rule:
% \begin{displaymath}
%   \inferrule*
%   {\Delta, i : s \vdash A \equiv B \isType}
%   {\Delta \vdash \forall i\mathord:s.A \equiv \forall i\mathord:s.B \isType}
% \end{displaymath}
% The congruence rules for the other type formers are similar.

\begin{lemma}
  Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
  substitution.
  \begin{enumerate}
  \item If $\Delta' \vdash e \equiv e' : s$ then $\Delta \vdash
    \sigma^*e \equiv \sigma^*e' : s$; and
  \item If $\Delta' \vdash A \equiv B \isType$ then $\Delta \vdash
    \sigma^*A \equiv \sigma^* B \isType$.
  \end{enumerate}
\end{lemma}

A pair of simultaneous substitutions $\Delta \vdash \sigma, \sigma'
\Rightarrow \Delta'$ are defined to be judgmentally equal if their
component expressions are judgmentally equal in the context $\Delta$:
i.e.~ if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
simultaneous substitutions are judgmentally equal.

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. Accordingly, we assume that each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretation of the
universal quantifier ignores the indexing: the interpretation of the
type $\forall i\mathord:s.A$ is exactly the interpretation of the type
$A$.

@@

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry instantiation of the general framework
  uses the assignment $\tyPrimSem{\tyPrimNm{vec}} = \mathbb{R}^2$ and
  $\tyPrimSem{\tyPrimNm{real}} = \mathbb{R}$.
\end{example}

% \begin{example}[Logical Information Flow]
%   \fixme{do this}
% \end{example}

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index expression equality: for
any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
substitutions $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash \sigma \equiv
\sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\
\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

@@

% The following lemma states that the relation interpretation of types
% that we have defined in this section behaves well: the first part of
% the lemma states that two types that are judgmentally equal are given
% equal relational interpretations, and the second part states that
% substitution of index expressions in types can be interpreted by the
% composition of relational environments with simultaneous
% substitutions.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and $\Delta \vdash \sigma
    \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}

Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.

@@

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework.

@@

The typing rules for our system define the well-typed programs with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgment $\Delta; \Gamma \vdash M : A$ is defined
in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a program that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}
